Composable data visualizations

### 2020-06-08 wip-compost/paper.tex
Add types and units section

\NeedsTeXFormat{LaTeX2e}
\documentclass{jfp}
\usepackage{amssymb}
\usepackage{xcolor}

@@

\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\fkvd}[1]{\textnormal{\textcolor{fkvdclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
\newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{\url{#1}}}}
\begin{document}
\maketitle[f]

@@

\begin{figure}[h]
  \includegraphics[scale=0.57]{figures/charts}
  \vspace{0.25em}
  \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
    and GBP-USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
  \label{fig:charts}
\end{figure}

@@

@@

Most charts have one X and one Y axis that are determined by the values the chart shows,
but there are interesting exceptions. The chart in Figure~\ref{fig:charts} (right)
has two different Y axes, one for GBP-USD and one for GBP-EUR. In the next two sections, we look at
three combinators that control the scale inferrence process and what flexibility this enables.

@@

The most interesting primitive for controlling scale composition defined in Figure~\ref{fig:control}
is $\kvd{nest}_{x/y}$. The combinator takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments
and it nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$. When inferring
scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
inferred using the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
the shape $s$. The chart space between $v_{max}$ and $v_{min}$ will then be used to render the
nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
Here, a chart with a continous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP-EUR exchange rates)
is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.

@@

In this paper, we focus on presenting the core ideas behind Compost. However, much
remains to be explored, both in terms of finding the best set of primitives and in terms of
their language integration. First, we only support categorical and continuous values, but
there are also ordinal values (which cannot be compared, but can be sorted). Second, some of our
primitives, namely \kvd{axis} and \kvd{roundScale} could be implemented as derived operations, but
we treat those as built-in for simplicity. Third, we only treat X and Y as scales, but we could
similarly treat other visual features (colors of bars, size of bubbles) as scales, which would
allow a more high-level specification of certain charts. Finally, there are also some obvious
ways in which a type system could prevent errors, for example by checking that categorical and
continuous values are not mixed on a single scale.

>>>>

\NeedsTeXFormat{LaTeX2e}
\documentclass{jfp}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{semantic}

@@

\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\fkvd}[1]{\textnormal{\textcolor{fkvdclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
\newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{\url{#1}}}}

\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}

\begin{document}
\maketitle[f]

@@

\begin{figure}[h]
  \includegraphics[scale=0.57]{figures/charts}
  \vspace{0.25em}
  \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
    and GBP/USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
  \label{fig:charts}
\end{figure}

@@

\subsection{Types and units of measure}
We introduce the Compost domain-specific language as untyped, but there are some obvious ways in
which types could make composing charts in Compost safer. First, a type representing a shape could
specify whether the X and Y axes represent categorical or continous values. This would rule out
mixing of different values on a single scale and guarantee that the \ident{union} operation,
sketched in the previous section, is never called in a way leading to an undefined result.
Second, the type of values mapped to an axis could be further annotated with units of measure
\cite{units}. Using the F\# notation where $n\langl u\rangl$ is a number $n$ with unit $u$,
an axis containing a value $\kvd{cont}~\num{317}\langl \ident{mp}\rangl$ would then be incompatible
with an axis containing a value $\kvd{cont}~\num{1.32}\langl \ident{gbp}/\ident{usd}\rangl$.

The design of the type system for Compost is straightforward, so we only present a brief sketch.
There are two kinds of types; $\sigma$ is a type of values and $\tau$ is a type of shapes.
Assuming $u$ denotes a unit of measure, the types are defined as:
%
\begin{equation*}
\begin{array}{rcl}
\sigma & = & \kvd{Cat}~u \lsep \kvd{Cont}~u
\end{array}
\qquad\qquad
\begin{array}{rcl}
\tau & = & \kvd{Shape}~\sigma_x, \sigma_y \\
\end{array}
\end{equation*}

\vspace{-1em}
\noindent
Correspondingly, there are two kinds of judgements; $v\vdash \sigma$ indicates the type of a
value, while $s\vdash \tau$ indicates the type of a shape. The typing rules for two of the
basic chart primitives, \kvd{line} and \kvd{overlay} look as follows:
%
\begin{equation*}
\inference
  {v_{xi} \vdash \sigma_x & v_{yi} \vdash \sigma_y}
  {\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] \vdash \kvd{Shape}~\sigma_x, \sigma_y}
\qquad
\inference
  {s_i \vdash \kvd{Shape}~\sigma_x, \sigma_y}
  {\kvd{overlay}~[\,s_1, \ldots, s_n\,] \vdash \kvd{Shape}~\sigma_x, \sigma_y}
\end{equation*}

\vspace{-0.5em}
\noindent
The rule for \kvd{line} ensures that all X and Y values have the same types, $\sigma_x$ and $\sigma_y$,
respectively and infers $\kvd{Shape}~\sigma_x, \sigma_y$ as the type of the shape. The rule for
\kvd{overlay} ensures that all composed shapes have the same type, including the type of X and Y scales.

@@

Most charts have one X and one Y axis that are determined by the values the chart shows,
but there are interesting exceptions. The chart in Figure~\ref{fig:charts} (right)
has two different Y axes, one for GBP/USD and one for GBP/EUR. In the next two sections, we look at
three combinators that control the scale inferrence process and what flexibility this enables.

@@

The most interesting primitive for controlling scale composition defined in Figure~\ref{fig:control}
is $\kvd{nest}_{x/y}$. The combinator takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments
and it nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$. When inferring
scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
inferred using the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
the shape $s$. The chart space between $v_{max}$ and $v_{min}$ will then be used to render the
nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
Here, a chart with a continous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP/EUR exchange rates)
is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.

@@

In this paper, we focus on presenting the core ideas behind Compost. However, much
remains to be explored, both in terms of finding the best set of primitives and in terms of
their language integration. First, we only support categorical and continuous values, but
there are also ordinal values (which cannot be compared, but can be sorted). Second, some of our
primitives, namely \kvd{axis} and \kvd{roundScale} could be implemented as derived operations, but
we treat those as built-in for simplicity. Third, we only treat X and Y as scales, but we could
similarly treat other visual features (colors of bars, size of bubbles) as scales, which would
allow a more high-level specification of certain charts.
